---
title: "Assignment_1_data_607"
author: "Jeff Shamp"
date: "1/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tracking Congress in the Age of Trump
## Introduction
538 updates a daily tally of who in congress supports Pres. Trump by voting in alingment with his agend. 
The also have built a prediction as to how frequently the Congressperson 'should' vote in alignment with Pres. Trump 
based on how that congressperson's district of state voted in the last general election. They also provide a simple
difference between how there actual record differs from the predicted record. 


I like this idea becuase it takes the general voting trend of a given district or state as a proxy for hwo aligned that 
location is with Pres. Trump's agend. It makes for an interesting view into how well a member of Congress aligns with
the voting record of their constituency vs. their alignment with Trump. 

For my analysis, I will be focusing on the Senate as they are in the spotlight right for the impeachment trial. 
I also think the Senate if significantly more out-of-touch with American than the House of Representatives. 

First we bring in the data and clean and process some values and columns
```{r}
library(RCurl)
d<-getURL("https://raw.githubusercontent.com/Shampjeff/cuny_msds/master/DATA_607/data/averages.csv")
d<-read.csv(text=d)
df<-data.frame(d)
senate_df<-subset(df, chamber=='senate')
senate_df$party[senate_df$party == "I"]<- "Independent"
senate_df$party[senate_df$party == "D"]<- "Democrat"
senate_df$party[senate_df$party == "R"]<-"Republican"
head(senate_df)
```
I'll also remove some of the columns that I do not think are important for this assignment. 

```{r}
senate_df<- subset(senate_df, select=-c(district,bioguide, chamber))
head(senate_df)
```
Now we have all the Senators, their party affliation, and voting stats relative to their constituency. 

The last thing that will be removed is the Congresspersons session-by-session record. I'll simply focus on the Senators
aggregated voting record for the time that Trump has been in office. There is a column title `congress` which lets the 
reader know during which session of Congress the votes were cast. I'll drop those rows and keep the combined record. 

```{r}
senate_df<- subset(senate_df, congress==0)
senate_df<- subset(senate_df, select=-c(congress))
head(senate_df)
```

## Exploratory Data Analysis
Now that we have cleaned and consistently formatted data, we can explore some basic questions. 

### Who votes for Trump when they do not have to?
Within this dataset is a column `net_trump_vote` which is the percent of the this consitutency that voted for Trump
minus the percent population that voted for Hillary Clinton. Negative values represent states that voted majority
for Clinton, and positive values are populations that majority voted for Trump. As such, it stands to reason that
low values, especially low negative values, represent populations that are swing states for Trump. I would not 
expect a Senator to vote very strongly pro-Trump is this regard, as they seem to be vunerable in elections. 

I'll look for Senators with 3% plus or minus `net_trump_vote` score that also have more than 60% vote alignment 
with the president. 

```{r}
swing_df<- subset(senate_df, agree_pct>.60 & net_trump_vote>-3 & net_trump_vote<=3)
swing_df
```
Indeed there is a very small number of Senators who vote well-aligned with Trump even though their home states are true swing states.

```{r}
plot(x=swing_df$net_trump_vote, y=swing_df$agree_pct, xlab='Net Trump Vote', ylab = 'Percent Alignment with Trump')
```

Above we see Susan Collins and Dean Heller (Heller no longer in Congress) having a
relatively high agreement with Trump despite representing states who voted for Clinton. The other on the right-hand side of the plot are Rick Scott, Marco Rubio, 
and Pat Toommey. With the exception of Collins these Senator are from classic 
swing states. 

If we look at each Senators predicted agreement with Trump vs. their actual agreement 
with Trump we see that the predicted values are not great overall. Perfect agreement
would be a linear relationship. 
```{r}
plot(senate_df$predicted_agree~senate_df$agree_pct, xlab='Percent Alignment with Trump',ylab = 'Predicted Alignment with Trump')
```

## Conclusions 
If I were a Democratic strategist, I would segment this information further to help identify
place to focus time and money to turn out Democratic voters. 

In terms of the predicted alignment with Trump, it needs more work. I would like to see primary 
voting data for swing state Senators like Rick Scott or Susan Collins as their alignment with 
Trump might be due to the sensibilities of Republican primary voters. 