---
title: "discussion week 13 - multi regression"
author: "Jeff Shamp"
date: "4/21/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Multiple Regression with Cars

This is the well-used well loved 1974 Motor Trend Cars dataset.

```{r}
data(mtcars)
head(mtcars)
```

We will plot a few variable and see what looks like a decent candidate for regression. 


```{r}
mtcars %>%
  ggplot(aes(x=mpg, y=hp)) +
  geom_point()
mtcars %>%
  ggplot(aes(x=mpg, y=wt)) +
  geom_point()
mtcars %>%
  ggplot(aes(x=mpg, y=disp)) +
  geom_point()
```

These are all fine options for quadratic regression on the surface. Since _there is no replacement for displacement_, I'll go with that variable. 

We will go with Transmission type (`am`) for the binary variable. 


```{r}
lm_1<- lm(data=mtcars,mpg ~ I(disp^2)+am+(disp*am))
summary(lm_1)
```

```{r}
lm_1 %>%
  ggplot(aes(sample = resid(lm_1))) +
  stat_qq() +
  stat_qq_line(color="red")

lm_1 %>%
  ggplot(aes(x=fitted(lm_1), y=resid(lm_1))) +
  geom_point() + 
  geom_smooth(method = "loess", color="red") +
  labs(x="predicted",
       y="residuals",
       title="Residual plot for predicted values")
```


Here the adjusted R square is okay but the error values are suspecious. Many are very close in value to the coefficient, which is not a good sign of good model. The quadratic term confers a positive association with mpg, as does the "manual" transmission categorical. Displacement and the interaction term of `am` and `disp` both have a neagtive association. The p-value for the interaction term is the highest of the three predictors, so that would be elminated if we wanted to refine this model. 

```{r}
lm_2<- lm(data=mtcars,mpg ~ I(disp^2)+am)
summary(lm_2)
```

```{r}
lm_2 %>%
  ggplot(aes(sample = resid(lm_2))) +
  stat_qq() +
  stat_qq_line(color="red")

lm_2 %>%
  ggplot(aes(x=fitted(lm_2), y=resid(lm_2))) +
  geom_point() + 
  geom_smooth(method = "loess", color="red") +
  labs(x="predicted",
       y="residuals",
       title="Residual plot for predicted values")
```


This appears to be a better model even if the Adjusted R square is not as high. The residuals look better and the p-values for the coefficients being non-zero are much better. 

