---
title: "discussion post hw2"
author: "Jeff Shamp"
date: "2/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set 2
## Matrix Factorization
This is from problem set 2 - I think I have a soluntion but I feel like it's a "hacky" soluntion and could be much tighter. 
Namely, is there a better way to handle the fact that the number of rows needed to elminate decrease with each iterative column. The first time around we need to elmination nrows-1 and column 1. The second time its nrows-2 and column 2. I use a row indexing variable `row_idx`, but I feel like this is inefficient. 

Write a function that factorizes any square matrix dimension 5x5 or less. No permuting rows necessary. 

```{r}
matrix.factorize<- function(input_mat){
  mat_L<-diag(dim(input_mat)[1])
  row_idx<-1
  for(j in 1:(dim(input_mat)[2]-1)){
    for(i in 1:(dim(input_mat)[1]-row_idx)){
      gcm<-(input_mat[i+row_idx,j]/input_mat[j,j])*(-1)
      augment_vec<- (gcm*input_mat[row_idx,])+input_mat[i+row_idx,]
      input_mat[i+row_idx,]<-augment_vec
      mat_L[i+row_idx,j]<-gcm*(-1)
    }
    row_idx<-row_idx+1
  }
  return(list(input_mat, mat_L))
}
```

Test the known example from the video posted in "Weekly Materials". 
```{r}
a<-matrix(c(2,4,-4,1,-4,3,-6,-9,5), byrow = T, nrow=3)
soln<-matrix.factorize(a)
soln[[2]]%*%soln[[1]]
```
Returns the input matrix under multiplication such that, $A = LU$. 

