---
title: "shamp_data_605_HW1"
author: "Jeff Shamp"
date: "1/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Set 1 
## General Approach
This is a straight forward problem set. "Size of one, size of the other, cosine of angle between them" is what my physics
undergraduate advisor use to say. We can take that approach here to solve most of the problems. 

### Problem 1 - Find the dot product of two vectors. 
```{r}
u<-c(0.5,0.5)
v<-c(3, -4)
dot_prod<-u%*%v
dot_prod
```
Easy enough. Make it a function.
```{r}
dot.product<- function(vec_1, vec_2){
  return(vec_1%*%vec_2)
}
```


### Problem 2- Vector length 
```{r}
vector.length<- function(vec_1) {
  dot_prod<-dot.product(vec_1, vec_1)
  return(sqrt(dot_prod))
}
```

We should test. The lenght of *v* should be 5. 

```{r}
v<-c(3,-4)
vector.length(v)
```
Great. 

### Problem 3 - Calculate 3u - 2v
```{r}
3*u - 2*v
```
It's that R is setup to use vectors and matricies so it feels like normal math expressions work as expected. 

### Problem 4
Vector dot products can be expressed as *u* ยบ *v* = ||*u*|| ||*v*|| *cos(X)*. Thus, dividing the dot product by the lenght of
the vectors will give use the cosine of the angle. 
```{r}
cosx<- (dot.product(u,v) / (vector.length(u)*vector.length(v)))
acos(cosx)
```
Above is the analytic answer when done by hand. 


# Set 2
## General Approach
The high school teacher in me wants to do this via the algebra two approach. So I'm going to build a function to solve any
system of two variables and simply run that function several times to reduce a system of three variables into two systems
of two variables. I can then progressively reduce the number of variables until the last variable has a solution.
Then use back-substitution to solve for the remaining variables. By hand, this is very time consuming, but it should lend itself
to automation fairly well. 

```{r}
two.equation.elim <- function(coeff_matrix, soln_arr) {
  # calculate the gcm for the first variable
  gcm<-coeff_matrix[1,1]*coeff_matrix[2,1]
  gcm_arr<-rep(NA, 2)
  gcm_arr[1]<-(gcm/coeff_matrix[1,1])
  gcm_arr[2]<-(gcm/coeff_matrix[2,1])
  
  # apply the gcm to each equation and constraint
  for(i in 1:2){
    coeff_matrix[i,]<-coeff_matrix[i,]*gcm_arr[i]
    soln_arr[i]<-soln_arr[i]*gcm_arr[i]
  } 
  #reduce the system by one variable
  reduce_eq<-coeff_matrix[1,]-(coeff_matrix[2,])
  reduce_soln<-soln_arr[1]-soln_arr[2]
  reduce_eq<-reduce_eq[-1]
  
  return(list(reduce_eq, reduce_soln))
}
```
Now I can write a function that takes in the matrix for a three variable system and its corresponding contraint array and solves 
for one variable. 

```{r}
three.variable.solver<- function(coeff_matrix, soln_arr) {
  if (dim(coeff_matrix)[1] == 3 & dim(coeff_matrix)[2] == 3){
    sys_1<-coeff_matrix[1:2,]
    sys_2<-coeff_matrix[2:3,]
    soln_1<-soln_arr[1:2]
    soln_2<-soln_arr[2:3]
  } else {
    print('Not a 3x3 matrix!')
  }
  sys_1_list<-two.equation.elim(sys_1, soln_1)
  sys_2_list<-two.equation.elim(sys_2, soln_2)
  
  result<- rep(NA, 3)
  #reshaping for maxtrix ops
  sys_3_coeff<-append(sys_1_list[[1]], sys_2_list[[1]])
  sys_3_soln<-append(sys_1_list[[2]], sys_2_list[[2]])
  sys_3_coeff<-matrix(sys_3_coeff, byrow=T, nrow=2)
  # final_var should be the solution to the third variable
  var_1<-two.equation.elim(sys_3_coeff, sys_3_soln)
  result[3]<- (var_1[[2]]/var_1[[1]])
  
  #sub back in and solve for the next variable
  y_values<-sys_1[,3]*result[3]
  soln_1<-soln_1-y_values
  var_2<-two.equation.elim(sys_1[,1:2], soln_1)
  result[2]<- (var_2[[2]]/var_2[[1]])
  
  x_values<-sys_1[,2]*result[2]
  soln_1<-soln_1-x_values
  result[1]<-(soln_1[1]/sys_1[1,1])
  return(result)
}
```

Let's test the known value
```{r}
test<-c(1,1,3,2,-1,5,-1,-2,4)
test_mat<-matrix(test, byrow = T, nrow = 3)
soln<- c(1,2,6)
three.variable.solver(test_mat, soln)
```

As excepted. Let's test for leading 0 coefficient.
```{r}
test<-c(1,1,3,-1,-2,4,0,3,3)
test_mat<-matrix(test, byrow = T, nrow = 3)
soln<- c(1,6,-2)
three.variable.solver(test_mat, soln)
```
So this is the zero pivot case, it seems. If the equation has a coefficient of 0 in the *x* position, then the system breaks
down. However, this situation is a simpler case to solve, so maybe a simple if/else can fix this issue. 




