---
title: "605 HW 15 - Jeff Shamp"
date: "`r Sys.Date()`"
output:
  pdf_document
---


```{r setup, message=FALSE}
library(ggplot2)
library(dplyr)
```



# Problem 1

**Find the equation of the regression line for the given points. Round any final values to the nearest hundredth, if necessary.**

( 5.6, 8.8 ), ( 6.3, 12.4 ), ( 7, 14.8 ), ( 7.7, 18.2 ), ( 8.4, 20.8 )

This can be done easily in R. 

```{r}
data<- as.data.frame(matrix(c(5.6,8.8, 6.3, 12.4, 7, 
                              14.8, 7.7, 18.2, 8.4, 20.8), 
                            byrow=TRUE, 
                            nrow = 5))
names(data)<- c("x", "y")

lin_reg<- lm(data=data, y ~ x)
summary(lin_reg)
```

```{r}
paste0("Equation of the line: y= ",
       lin_reg$coefficients[1],
       " + ", 
       round(lin_reg$coefficients[2],2),"*x")
```

```{r}
data %>%
  ggplot(aes(x=x, y=y)) +
  geom_point(color="red") + 
  geom_abline(slope = lin_reg$coefficients[2], 
              intercept = lin_reg$coefficients[1])
```

## Problem 2

**Find all local maxima, local minima, and saddle points for the function given below. Write your answer(s) in the form**

( x, y, z ). Separate multiple points with a comma.
$f(x,y) = 24x-6xy^{2}-8y^{3}$

To find the minimum we need the critical points for the partial derivatives. The partial derivatives are. 

$f(x,y)=24x-6xy^{2}-8y^{3}$

$f'_{x}(x,y)=\frac{d}{dx}(24x-6xy^{2}-8y^{3})=24-6y^{2}$

$f'_{y}(x,y)=\frac{d}{dy}(24x-6xy^{2}-8y^{3})= -12xy-24y^{2}$

Critical points are found by setting the partials to zero

$24-6y^{2}=0$

$y=\pm 2$

and

$-12xy-24y^{2}=0$ 

$x=-2y$

Thus we get the points, $(-4, 2), (4, -2)$ for the xy plane. We can plug these values into the original equation to obtain the 3-space solutions for the minimum. 

$24(-4)-6(-4)(2^{2})-8(2)^{3}= -64$

$24(4)6-(4)(-2^{2})-8(-2)^{3}=64$ 

So we have critical points of $(x,y,z) = (-4,2,-64), (4,-2,64)$

For saddle points we can to the second derivative test for $D(p)<0$ then p is a saddle point. 

$\frac{d^{2}}{dx}(24-6y^{2})=0$

$\frac{d^{2}}{dy}(-12xy-24y^{2})=-12x-48y$

$\frac{d^{2}}{dy}(24-6y^{2})=-12y$

Since $D(x,y)=f"_{x}(x,y)*f"_{y}(x,y)-f"_{xy}(x,y)$ and $f"_{x} = 0$ D reduces simply to $-(144)^{2}=-576$

So D(x,y) is less than zero, and both of these points are saddle points. 


# Problem 3

I'm not sure that part of this problem is calculus as it seems like there is an easy answer to this.

**A grocery store sells two brands of a product, the "house" brand and a "name" brand. The manager estimates that if she sells the "house" brand for x dollars and the "name" brand for y dollars, she will be able to sell**

**81 - 21x + 17y**

**units of the "house" brand and 40 + 11x - 23y units of the "name" brand.**

**Step 1. Find the revenue function R (x, y).**

Revenue is simply the unit sold multiplied by their price per unit. So $x(81- 21x + 17y)  + y(40 +11x -23y)$

Step 2. What is the revenue if she sells the "house" brand for \$2.30 and the "name" brand for \$4.10?

```{r}
x<-2.3
y<-4.1
R<-x*(81- 21*x + 17*y)  + y*(40 +11*x -23*y)
paste("Revenue =", R, "USD")
```


# Problem 4

**A company has a plant in Los Angeles and a plant in Denver. The firm is committed to produce a total of 96 units of a product each week. The total weekly cost is given by**

$C(x, y) = \frac{1}{6} x^{2} + \frac{1}{6} y^{2} + 7x + 25y + 700$, 

**where x is the number of units produced in Los Angeles and y is the number of units produced in Denver. How many units should be produced inveach plant to minimize the total weekly cost?**


Since we must produce 96 in total for the two plants we have, $96= x+y$ thus $y = 96-x$. If we substitute this into the original equation and do a bunch to arithmetic we get the following equation for cost, C(x,y). 

$C(x) = \frac{1}{3}x^{2} - 50x +4636$

Again if we wantt a minimum we can set the critical points to zero and solve. 

$C'(x)= \frac{2}{3}x -50 = 0$

so x = 75. Checking the second derivate for true minimum we get, $C"(x) = \frac{2}{3}$, which is greater than zero. 

Since 75 is the number of unit produced in LA, we know that 21 units must be produced in Denver to minimize cost. 


# Problem 5

**Evaluate the double integral on the given region.**

$\int_{2}^{4}\int_{2}^{4}e^{(8x+3y)}dA$

**Write your answer in exact form without decimals.**


This is easily separable, which is nice. 

$\int_{2}^{4}\int_{2}^{4}e^{(8x+3y)}dxdy$

$\int_{2}^{4}e^{(8x)}dx*\int_{2}^{4}e^{(3y)}dy$

$\frac{1}{24}e^{8x} | ^{4}_{2}*e^{3y} | ^{4}_{2}$

So in exact form we have

$A = \frac{1}{24}(e^{32}-e^{16})(e^{12}-e^{6})$

Which is some astronomical number. 






