---
title: "HW 8 Data 605"
author: "Jeff Shamp"
date: "3/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 11 Page 303

In general the expeccted value for exponential distributions are as follows:

$E\left [ X_{i} \right ]= \frac{1}{\lambda_{i}}$. 

For $X_{i}$ independent random variables we know that the expected value is simply:

$E\left [ X_{i} \right ]= \frac{1}{1000}$

If we are looking for the _first_ lightbuld to burn out then we are interested in the minimum of the $X_{i}$ from n=1 to n=100. For each lightbulb we have an expected value of $\frac{1}{1000}$ and using the ideas of CDF we can sum those values across all members of $X_{i}$ to find the minimum expected value of $X_{i}$ Resulting in:

$\sum_{1}^{100}\lambda_{i}=\frac{100}{1000}=\frac{1}{10}$

Thus, the minimum expected value is;

$min(E\left [ X_{i} \right ])= \frac{1}{\frac{1}{10}}=10$. We can expect the first lightbulb to burn out in 10 hours. 

## Problem 14 page 303

This is a convolution problem for a difference of two random variable with the same exponential density, $f_{z}(z) = \lambda e^{(-\lambda z)}$

Using the PDF ideas from the video we see that this be treated similarly, 

$\int_{-\infty }^{\infty}f_{x1}(z-x) f_{x2}(x_{1})dx_{2}$


Given that we already know $f_{x1}$ and $f_{x2}$ are the exponential density functions we can rewritte the above as the following.

$\int_{-\infty }^{\infty}\lambda^{2}e^{-\lambda(z+2x_{x2})}e^{\lambda z}dx_{1}$

As much as I love to do this integral by hand, the internet has told me that this nicely reduces to the following states;

$f_{z}(z) =\left\{\begin{matrix}
\frac{\lambda}{2} e^{\lambda z},  z<0 \\ 
\frac{\lambda}{2} e^{-\lambda z},  z>0
\end{matrix}\right.$

This defines the desired outcome $\frac{\lambda}{2} e^{-\lambda \left | z \right |}$ *Note that $\lambda$ is negative for the positive z.


## Problem 1 page 320

Let X be a continuous random variable with mean µ = 10 and variance
$\sigma^{2} = 100/3$. Using Chebyshev’s Inequality, find an upper bound for the
following probabilities.

a. $P(|X-10| \geq 2)$

```{r}
var <- 100/3
k_square<-4
upper<- var/k_square
upper
```

Since this is a probability, we take it to be 1.

b. $P(|X - 10|) \geq 5)$

```{r}
k_square<-25
upper<- var/k_square
upper
```

Again, this is a probability so we take it to be 1.

c. $P(|X-10| \geq 9)$

```{r}
k_square<-81
upper<- var/k_square
upper
```

Upper bound is around 40%.


d. $P(|X-10|\geq 20)$

```{r}
k_square<-400
upper<- var/k_square
upper
```

Upper bound is about 8%

