title: "Data 605 HW 5"
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="75")
opts_knit$set(width=75)
```

## Problem Set 1

<br>
Choose independently two numbers B and C at random from the interval [0, 1] with
uniform density. Prove that B and C are proper probability distributions.
Note that the point (B,C) is then chosen at random in the unit square.

For this first task I am going to build a simulation in R and run thousands of trials rather than write out integrals in Latex. 

```{r}
B<- runif(10000, min=0, max=1)
C<- runif(10000, min= 0, max=1)
```

This should give us what we want. If you do not trust your machine overloads, then we can examine the data a bit to satisfy ourselves B and C are $[0,1]$. Since C and B are created the same way, if it works for B, then it should work for C as well. 

```{r}
B_0<-sum(B < 0)
B_1<-sum(B > 1)
print(paste("number of values less than 0:", B_0))
print(paste("number of values greater than 1:", B_1))
```

Also we can sample these vectors easily.
```{r}
sample(C, size = 20)
```

I'm convinced, but does anyone have a better way to compare these values with eye-balling it? Or a more rigorous way to show that B and C are "proper probability distributions" ?
