---
title: "jshamp assignment2 PS1 and PS2"
author: "Jeff Shamp"
date: "2/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem Set 1

## Problem 1 - Prove $A^{T}A\neq AA^{T}$

### Proof by induction:
**Assume:** $A^{T}A = AA^{T}$ 

Then for rows *i* in A and columns *j* in $A^{T}$, $A_{i1}\cdot A^{T}_{1j} = A^{T}_{i1}\cdot A_{1j}$. This is true if and only if $A_{i1} = A^{T}_{1j}$, such that $A = A^{T}$. Therefore, the assumption is false outside of the rare case of identical matricies. Thus, $A^{T}A\neq AA^{T}$. 

### Example:
Consider the following; $A =\begin{bmatrix}1 & 1 & 1\\ 2 &2 & 2 \\ 3 & 3 & 3\end{bmatrix}$ and $A^{T} =\begin{bmatrix} 1 & 2 & 3\\ 1 &2 & 3 \\ 1 & 2 & 3\end{bmatrix}$

Here if we take the inner product of the first columns of $A$ and the first row of $A^{T}$ we get $A_{i1} = 14$. Similarly, we take the first inner product of $A^{T}$ and $A$ we get $A^{T}_{1j} = 3$, such at $A_{i1} \neq A^{T}_{1j}$ and $A^{T}A\neq AA^{T}$. 


## Problem 2 -Under what circumstances could Problem 1 be true?

As mentioned in the proof, this is true for any matrix $A$ that is elementwise identical to its transpose $A^{T}$. That is, $A = A^{T}$. 

___
\pagebreak

# Problem Set 2
## Matrix Factorization
Write a function that factorizes any square matrix less the dimension 5x5. No permuting rows necessary. 

```{r}
matrix.factorize<- function(input_mat){
  mat_L<-diag(dim(input_mat)[1])
  row_idx<-1
  for(j in 1:(dim(input_mat)[2]-1)){
    for(i in 1:(dim(input_mat)[1]-row_idx)){
      gcm<-(input_mat[i+row_idx,j]/input_mat[j,j])*(-1)
      augment_vec<- (gcm*input_mat[row_idx,])+input_mat[i+row_idx,]
      input_mat[i+row_idx,]<-augment_vec
      mat_L[i+row_idx,j]<-gcm*(-1)
    }
    row_idx<-row_idx+1
  }
  return(list(input_mat, mat_L))
}
```

Test the known example from the video posted in "Weekly Materials". 
```{r}
a<-matrix(c(2,4,-4,1,-4,3,-6,-9,5), byrow = T, nrow=3)
soln<-matrix.factorize(a)
soln[[2]]%*%soln[[1]]
```
Returns the input matrix under multiplication such that, $A = LU$. 

Let's try one more from the video. 
```{r}
b<-matrix(c(1,4,-3,-2,8,5,3,4,7), byrow = T, nrow=3)
soln_2<-matrix.factorize(b)
soln_2[[2]]%*%soln_2[[1]]
```
Solid. 


