---
title: "Data 605 - HW3"
author: "Jeff Shamp"
date: "2/11/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 1

1) Rank of given matrix $A$. For a square matrix the rank will be equal to the number of rows so long as the Upper Triangular ( I'll call $U$) form of the matrix has a non-zero determinant. If det($U$) = 0, then we know the rank is less than the dimension and will need to repeat this process for N-1 dimension submatrix until a non-zero determinant is found. For some cases, visual inspection might be better but those are rare cases and do not scale well. 

I'm going to bring in a function from last weeks HW to calculate the Upper Triangular matrix. 

```{r}
matrix.factorize<- function(input_mat){
  mat_L<-diag(dim(input_mat)[1])
  row_idx<-1
  for(j in 1:(dim(input_mat)[2]-1)){
    for(i in 1:(dim(input_mat)[1]-row_idx)){
      mat_L[i+row_idx,j]<-(input_mat[i+row_idx,j] /
                             input_mat[j,j])
      input_mat[i+row_idx,]<-((-1*mat_L[i+row_idx,j]) *
                                input_mat[row_idx,]) +
                                input_mat[i+row_idx,]
      }
  row_idx<-row_idx+1
    }
  
  return(list(input_mat, mat_L))
}
```

```{r}
A<- matrix(c(1,2,3,4,-1,0,1,3,0,1,-2,1,5,4,-2,-3), byrow=T, nrow=4)
soln<-matrix.factorize(A)
det(soln[[1]])
```

Thus the rank of matrix $A$ is 4, since it has a non=zero determinant. I'm going to try another example where I know the answer, just to be sure this works. A 4x4 checkerboard matrix of 1s and -1s should have a rank of 1.

```{r}
C<-matrix(c(1,-1,1,-1,-1,1,-1,1,1,-1,1,-1,-1,1,-1,1), byrow = T, nrow = 4)
soln<- matrix.factorize(C)
det(soln[[1]])
```

This is the expected answer for this matrix. Let's take one step further and find the rank of square matrix. I'll return a vector of all the determinants as well for redundacy and transparency. 

```{r}
find.matrix.rank<- function(mat){
  mat_dim<-dim(mat)[1]-1
  rank<-dim(mat)[1]
  mat_det_vec<-rep(NA,mat_dim)
  for (i in 1:mat_dim){
    mat_det_vec[i]<-det(matrix.factorize(mat)[[1]])
    if (mat_det_vec[i]!=0){
      return(list(rank, mat_det_vec))
  } else {
    rank<-rank-1
    mat<-mat[1:dim(mat)[1]-1, 1:dim(mat)[2]-1]
  }
  }
  return(list(rank, mat_det_vec))
}
find.matrix.rank(C)
```

As expected. Let's re-try problem 1.1 with the same matrix $A$.

```{r}
find.matrix.rank(A)
```
Excellent. 

2) The minimum rank for any non-zero matrix is rank =1. The maximum is no greater than the lesser of the dimensions. In this case, *n*. 

3) Rank of a given matrix can be found with out handy function from 1.1

```{r}
B<-matrix(c(1,2,1,3,6,3,2,4,2), byrow=T, nrow=3)
find.matrix.rank(B)
```

This makes sense as this represents three parallel lines, which is a dependent system. 

## Problem 2 - Eigenvectors and values

We are to calculate the eigenvector and values for a 3x3 matrix in Upper Triangular form, which is helpful. I'll do most of this work by-hand but with some machine computation. First set up the $det(A - \lambda I)$ equation. 

$det(\begin{bmatrix}
1 & 2 & 3\\ 
0 & 4 &5 \\ 
0 & 0 & 6
\end{bmatrix} - \begin{bmatrix}
\lambda  &0  &0 \\ 
0 & \lambda &0 \\ 
 0& 0 & \lambda 
\end{bmatrix}) =0$

Which expands to the following:

$(1-\lambda)det(\begin{bmatrix}
4-\lambda & 5 \\ 
0 & 6-\lambda  
\end{bmatrix} ) - 
(2)det(\begin{bmatrix}
0 &5 \\ 
0 & 6-\lambda
\end{bmatrix}) +
(3)det(\begin{bmatrix}
0 &4-\lambda \\ 
0 & 0
\end{bmatrix}) =0$

Thankfully, the last two terms of this expansion reduce to zero. Thus we are left with the following:

$(1 - \lambda)(24-10 \lambda + \lambda^{2}) =0$

This expands to the following,

$24-34\lambda+11\lambda^{2}-\lambda^{3}=0$

Which can be factored via the fundamental theorem of algebra (luckily $\lambda=1$ is a root!). We see that this reduces to, 

$-\lambda^{2}+10\lambda-24 =0$. Which factors our last two remaining eignevalues and our set of eigenvalues are as follows:

$\lambda_{1}=1, \lambda_{2}=4, and \lambda_{3}=6$.

If we return the first eigenvalue in the original matrix $A$ we apply $\lambda*I - A$ as the following,

$\begin{bmatrix}
0 & 2 &3 \\ 
 0&3  &5 \\ 
0 &0  &5 
\end{bmatrix} \begin{bmatrix}
v_{1}\\ 
v_{2}\\ 
v_{3}
\end{bmatrix}=0$

The second eigenvalue can be used to obtain the eigenvector in the same way subtract $\lambda = 4$ from the diagonal of the original matrix $A$ and the same is true for the the last eigenvalue. 

$\begin{bmatrix}
-3 & 2 &3 \\ 
 0& 0  &5 \\ 
0 &0  & 2 
\end{bmatrix} \begin{bmatrix}
v_{1}\\ 
v_{2}\\ 
v_{3}
\end{bmatrix}=0$

R can compute the values of the eigenvectors and reduce them to unit values by calling the `eigen(A)$vectors` command. 

```{r}
A<-matrix(c(1,2,3,0,4,5,0,0,6), byrow = T, nrow = 3)
eigen(A)$vectors
```

This is the span of vectors for the Eigenspace that are characteristic to martix $A$. 











