---
title: "606 presentation - probability model"
author: "Jeff Shamp"
date: "2/27/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Problem - Create A Probability Model

**4.35 Roulette winnings.** In the game of roulette, a wheel is spun and you place bets on where it will stop. One popular bet is that it will stop on a red slot; such a bet has an 18/38 chance of winning. If it stops on red, you double the money you bet. If not, you lose the money you bet. Suppose you play 3 times, each time with a $1 bet. Let Y represent the total amount won or lost. Write a probability model for Y.

- Binomial $\binom{n}{k}(p)^{k}(1-p)^{n-k}$

## Ways To Win

- 0 wins:  Y = -3 <br>
  $\binom{3}{0}(\frac{18}{38})^{0}(\frac{20}{38})^{3} = 14.5\%$
  
- 1 win: Y= 0  - break even! <br>
  $\binom{3}{1}(\frac{18}{38})^{1}(\frac{20}{38})^{2} = 39.3\%$
  
- 2 wins: Y = 3 <br>
  $\binom{3}{2}(\frac{18}{38})^{2}(\frac{20}{38})^{1} = 35.4\%$
  
- 3 wins: Y= 6  - The least likely event <br>
  $\binom{3}{3}(\frac{18}{38})^{3}(\frac{20}{38})^{0} = 10.6\%$

## Distribution

```{r cars, echo = TRUE}
x<-dbinom(0:3, size=3, prob=(18/38)) 
barplot(x, 1:3)
```


