---
title: "Data 607 HW3"
author: "Jeff Shamp"
date: "2/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem #1
538 dataset on college majors. Identify the majors with either data or statistics in the title. 

```{r, results='hide'}
library(stringr)
library(RCurl)
```

```{r}
d<-getURL("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv")
df<-data.frame(read.csv(text=d))
head(df)
```

```{r}
str_subset(df$Major, pattern="DATA|STATISTICS")
```

That was a lot easier than I thought it would be. 


## Problem #2 - Vectorize long fruits string


```{r}
data<-paste('[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"')
```

```{r}
data
```

Yikes. Let's drop the `\` with `[ a-z]` and match it at least once with `+`

```{r}
str_view(data, "[ a-z]+")
```

Now we can unlist and anchor with an empty string to get the spaces. 

```{r}
fruits<-c(unlist(str_extract_all(data, "\\b[ a-z]+\\b")))
fruits[5:9]
```

```{r}
typeof(fruits)
```

So `fruits` is now a character vector. Nice. 

## Problem #3 - R for Data Science textbook problems

### Describe, in words, what these expressions will match:

- (.)\1\1

**This is a syntax error. There are no quotes. If it were, "(.)\1\1", it would match a \1 in the string.**

- "(.)(.)\\\2\\\1"

**This creates two 'wildcard' characters and matches a repeat for the second item followed by a repeat of the first. **
**Example: in 'xloolz' this command would match 'lool'**

- (..)\1

**This is another syntax error. Again, no quotes. If there were quotes it would match any character that is in front of a \1. Example "That store has \1 a banh mi station!", would match the 's' in 'has' followed by the space.**

- "(.).\\\1.\\\1"

**The creates a 'wildcard' item and any following character and matches the item again. It then matches any second character and repeats the original item a third time.**
**Example: in the string, 'basaxa' this command would match, 'asaxa' The item is 'a', any character, 'a', any character, 'a'.**

- "(.)(.)(.).*\\\3\\\2\\\1"

**This looking for any string that has three characters in a given order and in reseverse order.**
**Example: In "abc123098cba" this command would select the entire string. In the string "abc123cba987" it would select "abc123cba" only.**

## Problem #4 - 
### Construct regular expressions to match words that:

- Start and end with the same character.

  -" ^(.).*\\\1$ "

- Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)

  -" (..).\*\\\1* "

- Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

  -" .\*(.).\*\\\1.\*\\\1* "



